---
title: "Clean fasta headers in the proteome files to be used in OrthoFinder analysis"
author: "Neha Sahu"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE, 
  warning = FALSE,
  fig.width = 10, 
  fig.height = 7
)
```

# Introduction

This document explains the script designed to handle raw fasta files with complex headers, non-uniform and file names, making them more user-friendly and easier to work with.

# Load libraries

```{r }
library(here)
library(tidyverse)
library(phylotools)
library(seqinr)
```

# List and read the raw fasta files

```{r }
file_names <- list.files(here("data/raw/sequences/","input_files"), full.names = TRUE)
file_names
```

# Read and merge into single data frame

```{r}
fasta_data <- map_dfr(file_names, ~ cbind(phylotools::read.fasta(.x), filename = basename(.x))) 
head(fasta_data)
```

# Add new columns for new names and new fasta headers that are simplified for further analyses

```{r}
fasta_data <- fasta_data %>%
  group_by(filename) %>%
  mutate(new_headers = paste0(substr(filename, 1, 5),"_",sub("^.*?_(.{5}).*", "\\1", filename), "_", row_number()), # Create a new_headers columns - abbreviates based on filename
         names=paste0(substr(filename, 1, 5),"_",sub("^.*?_(.{5}).*", "\\1", filename))) # For replacing complicated filenames with simple names
length(unique(fasta_data$names))
length(unique(fasta_data$filename))
#The above two numbers should be the same
fasta_data$seq.text <- gsub("[^A-Za-z0-9]", "", fasta_data$seq.text) # Remove special characters with nothing
```

# Write each split data frame to a separate fasta file which will be saved in the fasta_files folder

```{r}
# Create directory to save the fasta files with cleaned headers
if (!dir.exists(here("data", "processed", "fasta_files"))) {
  dir.create(here("data", "processed", "fasta_files"), recursive = TRUE)
}

# Split the fasta data by names
fasta_data_split <- split(fasta_data, f = fasta_data$names)

# Write each species to a separate fasta file
result <- lapply(names(fasta_data_split), function(x) {
    # Create the full path using here() to be consistent
    output_file <- here("data", "processed", "fasta_files", 
                        paste0(fasta_data_split[[x]]$names[1], ".fasta"))
    # Write fasta files
    write.fasta(as.list(fasta_data_split[[x]]$seq.text), 
                fasta_data_split[[x]]$new_headers, 
                file.out = output_file,  # Use the full path with here()
                open = "w", 
                as.string = FALSE)
})
```

# IMPORTANT

**BEFORE RUNNING OrthoFinder: To this final folder called fasta_files, replace the Magna_oryzae.fasta with MG8 genome from Ensembl (**<https://fungi.ensembl.org/Magnaporthe_oryzae/Info/Index>) **to have a proteome with compatible MGG_IDs**

```{r}
sessionInfo()
```
