---
title: "Check if DEGs in Dbip1 and Dmst12 are co-regulated or co expressed"
author: "Neha Sahu"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  github_document:
    toc: true
    toc_depth: 3
    dev: "png"
    html_preview: false
  html_document:
    toc: true
    toc_float: true
    theme: yeti
    highlight: pygments
    code_folding: show
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
library(here)
source(here("scripts","analysis_config.R"))
init_analysis("003_co_expression")

knitr::opts_chunk$set(
  fig.path="../data/results/003_co_expression/00_figures/"
)
```

# Description

This analysis continues from 002_rna_seq.Rmd and checks the
co-expression patterns between mst12 and bip1 genes in *Magnaporthe
oryzae*. The goal is to see if bip1 and mst12 are co-regulated by
checking their expression patterns across different mutant strains
(Dbip1, Dmst12) and timepoints (0h, 4h, and 24h). In simple terms, I
want to see if mst12 and bip1 show coordinated/similar regulation
patterns.

## Objectives:

1.  Create Euler diagrams to visualize overlap of differentially
    expressed genes between mutant strains

2.  Generate correlation matrices to assess co-expression patterns

3.  Make heatmaps to visualize expression profiles of differently
    expressed genes

## Load Libraries

```{r load_lib, message=F, warning=F}
library(tidyverse)
library(eulerr) # for eulerr plot
library(corrplot) # for correlation plot
library(gplots) # for heatmap
```

# **Eulerr plots**

### Merged DEGs File

```{r merge_degs, message=F, warning=F}
merged_data <- read_csv(here("data/results/002_rna_seq/merged_DEGs_guy11_vs_Dbip1_Dmst12.csv"))

degs_filt <- merged_data %>%
  select(target_id,Comparison,DEG_status) %>%
  filter(DEG_status!="not_DEG") %>%
  mutate(time=word(Comparison,sep="_",4,4),
         sample=word(Comparison,sep="_",3,3),
         DEG_status=gsub("_in_guy11","",DEG_status)) %>%
  select(-Comparison) %>%
  distinct()
unique(degs_filt$time)


```

### Downregulated in Guy11 - Upregulated in Dbip1, Dmst12

```{r downregs_in_guy11_euler, fig.align='center',fig.width=10,fig.height=7, warning=F}
# Function to create Euler plots for each timepoint
create_euler_plot <- function(time_point, deg_status, title_suffix) {
  downreg <- degs_filt %>%
    filter(DEG_status == deg_status, time == time_point) %>%
    distinct() %>%
    select(-DEG_status, -time) %>%
    arrange(target_id, factor(sample, levels = c("Dbip1", "Dmst12"))) %>%
    aggregate(. ~ target_id, toString)
  
  downreg_set <- data.frame(table(downreg$sample))
  downreg_set$Var1 <- gsub(", ", "&", downreg_set$Var1)
  downreg_data <- setNames(downreg_set$Freq, downreg_set$Var1)
  
  euler_plot <- euler(downreg_data, shape = "ellipse", input = c("disjoint"), loss = c("abs"))
  
  # Save data
  save_data(downreg_set, paste0("euler_", gsub(" ", "_", tolower(title_suffix)), "_", time_point))
 
  return(list(data = downreg_set, euler_obj = euler_plot))
}

# Create plots for downregulated genes at each timepoint
downreg_0h <- create_euler_plot("0h", "downreg", "Downregulated genes in guy11")
downreg_4h <- create_euler_plot("4h", "downreg", "Downregulated genes in guy11")
downreg_24h <- create_euler_plot("24h", "downreg", "Downregulated genes in guy11")

plot(downreg_0h$euler_obj,fills = list(fill = c("#fdc086", "#beaed4", "#ffff99")),
     legend = TRUE,
     quantities = TRUE,
     labels = list(font = 4, cex = 1),
     lty = 1:4,
     main = "Downregulated genes in guy11 at 0h")
plot(downreg_4h$euler_obj,fills = list(fill = c("#fdc086", "#beaed4", "#ffff99")),
     legend = TRUE,
     quantities = TRUE,
     labels = list(font = 4, cex = 1),
     lty = 1:4,
     main = "Downregulated genes in guy11 at 4h")
plot(downreg_24h$euler_obj,fills = list(fill = c("#fdc086", "#beaed4", "#ffff99")),
     legend = TRUE,
     quantities = TRUE,
     labels = list(font = 4, cex = 1),
     lty = 1:4,
     main = "Downregulated genes in guy11 at 24h")

```

### Upregulated in Guy11 - Downregulated in Dbip1, Dmst12

```{r upregs_in_guy11_euler, fig.align='center',fig.width=10,fig.height=7, warning=F}
# Function to create Euler plots for each timepoint
create_euler_plot <- function(time_point, deg_status, title_suffix) {
  upreg <- degs_filt %>%
    filter(DEG_status == deg_status, time == time_point) %>%
    distinct() %>%
    select(-DEG_status, -time) %>%
    arrange(target_id, factor(sample, levels = c("Dbip1", "Dmst12"))) %>%
    aggregate(. ~ target_id, toString)
  
  upreg_set <- data.frame(table(upreg$sample))
  upreg_set$Var1 <- gsub(", ", "&", upreg_set$Var1)
  upreg_data <- setNames(upreg_set$Freq, upreg_set$Var1)
  
  euler_plot <- euler(upreg_data, shape = "ellipse", input = c("disjoint"), loss = c("abs"))
  
  # Save data and plot
  save_data(upreg_set, paste0("euler_", gsub(" ", "_", tolower(title_suffix)), "_", time_point))
  
  return(list(data = upreg_set, euler_obj = euler_plot))
}

# Create plots for upregulated genes at each timepoint
upreg_0h <- create_euler_plot("0h", "upreg", "upregulated genes in guy11")
upreg_4h <- create_euler_plot("4h", "upreg", "upregulated genes in guy11")
upreg_24h <- create_euler_plot("24h", "upreg", "upregulated genes in guy11")

plot(upreg_0h$euler_obj,fills = list(fill = c("#fdc086", "#beaed4", "#ffff99")),
     legend = TRUE,
     quantities = TRUE,
     labels = list(font = 4, cex = 1),
     lty = 1:4,
     main = "Upregulated genes in guy11 at 0h")
plot(upreg_4h$euler_obj,fills = list(fill = c("#fdc086", "#beaed4", "#ffff99")),
     legend = TRUE,
     quantities = TRUE,
     labels = list(font = 4, cex = 1),
     lty = 1:4,
     main = "Upregulated genes in guy11 at 4h")
plot(upreg_24h$euler_obj,fills = list(fill = c("#fdc086", "#beaed4", "#ffff99")),
     legend = TRUE,
     quantities = TRUE,
     labels = list(font = 4, cex = 1),
     lty = 1:4,
     main = "Upregulated genes in guy11 at 24h")

```

# **Correlation Between Samples**

### Load TPM table

```{r load_tpm}
tpm <- read_csv(here("data/results/002_rna_seq/TPM_table.csv"))
colnames(tpm)

tpm_corr<-data.matrix(tpm[,-1])
head(tpm_corr[1,])
```

### Calculate and Plot Pearson Correlation Between Samples - For All Genes

```{r plot_pcorr, fig.align='center',fig.width=10,fig.height=7, warning=F }
M<-cor(tpm_corr, method="pearson",use="pairwise.complete.obs")
head(M[1,])
corrplot(M,type="full",method="circle",
         cl.cex=0.5,
         main="All expressed genes", is.corr=F,col = COL2('BrBG'),
         tl.cex = 1,tl.srt = 45,addCoef.col = 'black',
         diag=T,
         tl.col = c(rep("#fdc086",3),#bip1
                    rep("#beaed4",3),#mst12
                    rep("#7fc97f",3))) #guy11
```

### Calculate and Plot Pearson Correlation Between Samples - For Genes Upregulated in Guy11

```{r plot_pcorr_upregs_guy11, fig.align='center',fig.width=10,fig.height=7, warning=F}

degs_pcorr <- merged_data %>%
  select(target_id,Comparison,DEG_status) %>%
  filter(DEG_status=="upreg_in_guy11") %>%
  mutate(time=word(Comparison,sep="_",4,4),
         sample=word(Comparison,sep="_",3,3)) %>%
  select(-Comparison) %>%
  filter(sample!="tbip1") %>%
  select(target_id) %>%
  distinct()
head(degs_pcorr)

tpm_degs <- tpm %>%
  filter(target_id %in% degs_pcorr$target_id) %>%
  select(target_id,
         Guy11_0h,Guy11_4h,Guy11_24h,
         Dbip1_0h,Dbip1_4h,Dbip1_24h,
         Dmst12_0h,Dmst12_4h,Dmst12_24h)
tpm_degs_corr<-data.matrix(tpm_degs[,-1])
head(tpm_degs_corr[1,])
M_degs<-cor(tpm_degs_corr, method="pearson",use="pairwise.complete.obs")
head(M_degs[1,])
corrplot(M_degs,type="full",method="circle",
         cl.cex=0.5,
         main="Upregs in Guy11", is.corr=F,col = COL2('BrBG'),
         tl.cex = 1,tl.srt = 45,addCoef.col = 'black',
         diag=T,
         tl.col = c(rep("#7fc97f",3),#guy11
                    rep("#fdc086",3),#bip1
                    rep("#beaed4",3)#mst12
                    )) 

```

### Calculate and Plot Pearson Correlation Between Samples - For Genes Downregulated in Guy11

```{r plot_pcorr_downregs_guy11, fig.align='center',fig.width=10,fig.height=7, warning=F}

degs_pcorr <- merged_data %>%
  select(target_id,Comparison,DEG_status) %>%
  filter(DEG_status=="downreg_in_guy11") %>%
  mutate(time=word(Comparison,sep="_",4,4),
         sample=word(Comparison,sep="_",3,3)) %>%
  select(-Comparison) %>%
  filter(sample!="tbip1") %>%
  select(target_id) %>%
  distinct()
head(degs_pcorr)

tpm_degs <- tpm %>%
  filter(target_id %in% degs_pcorr$target_id) %>%
  select(target_id,
         Guy11_0h,Guy11_4h,Guy11_24h,
         Dbip1_0h,Dbip1_4h,Dbip1_24h,
         Dmst12_0h,Dmst12_4h,Dmst12_24h)
tpm_degs_corr<-data.matrix(tpm_degs[,-1])
head(tpm_degs_corr[1,])
M_degs<-cor(tpm_degs_corr, method="pearson",use="pairwise.complete.obs")
head(M_degs[1,])
corrplot(M_degs,type="full",method="circle",
         cl.cex=0.5,
         main="Downregs in Guy11", is.corr=F,col = COL2('BrBG'),
         tl.cex = 1,tl.srt = 45,addCoef.col = 'black',
         diag=T,
         tl.col = c(rep("#7fc97f",3),#guy11
                    rep("#fdc086",3),#bip1
                    rep("#beaed4",3)#mst12
                    )) 


```

# **Heatmaps**

### Load DEGs in Dmst12 and Dbip1 for Heatmap

```{r load_degs_hm}
degs_heat <- merged_data %>%
  select(target_id,Comparison,DEG_status) %>%
  filter(DEG_status!="Not_DEG") %>%
  mutate(time=word(Comparison,sep="_",4,4),
         sample=word(Comparison,sep="_",3,3)) %>%
  select(-Comparison) %>%
  filter(sample!="tbip1") %>%
  mutate(DEG_status=gsub("_in_guy11","",DEG_status)) %>%
  left_join(tpm,by="target_id") %>%
  distinct()
head(degs_heat)
```

### Upregs in Guy11

```{r upreg_heatmap, fig.align='center',fig.width=7,fig.height=15, warning=F}
upreg<-degs_heat %>%
  filter(DEG_status=="upreg") %>%
  select(-DEG_status,-time,-sample) %>%
  distinct() %>%
  column_to_rownames(var="target_id") %>%
  select(Guy11_0h, Guy11_4h, Guy11_24h,
         Dbip1_0h,Dbip1_4h,Dbip1_24h,
         Dmst12_0h,Dmst12_4h,Dmst12_24h)

upreg_matrix<-data.matrix(upreg)

# Use hierarchical clustering
mycols = colorRampPalette(c("blue","white","red"))(100) 
hr <- hclust(as.dist(1-cor(t(upreg_matrix), method="pearson")), method="average") 
dend <- as.dendrogram(hr) # Convert to dendrogram
row_order <- order.dendrogram(dend) # Row order from the hierarchical clustering
upreg_matrix_ordered <- upreg_matrix[row_order, ] %>%  # Reorder matrix based on the clustering
  as.data.frame() %>%
  rownames_to_column(var = "target_id")  # Add row names as a column

# Save ordered matrix
save_data(upreg_matrix_ordered, "upreg_matrix_ordered", format = "txt")

pdf_path <- here::here("data/results/00_figures/003_co_expression", "upregs_guy11vsDbip1_Dmst12.pdf")
#pdf(pdf_path, width = 12, height = 20)
heatmap.2(upreg_matrix, 
                  Rowv = as.dendrogram(hr), 
                  dendrogram = "row", 
                  trace = "none",  
                  col = mycols, 
                  keysize = 1, 
                  key.title = "Title", 
                  Colv = FALSE,
                  scale = c("row"), 
                  cexRow = 0.5, 
                  cexCol = 1, 
                  srtCol = NULL,  
                  srtRow = NULL, 
                  adjRow = c(0,NA), 
                  sepcolor = "white", 
                  margins = c(7,15), 
                  lwid = c(2,8),
                  lhei = c(2,8), 
                  main = "upregs", 
                  na.color = "black")
```

### Downregs in Guy11

```{r downreg_heatmap, fig.align='center',fig.width=7,fig.height=15, warning=F}
downreg<-degs_heat %>%
  filter(DEG_status=="downreg", sample!="tbip1") %>%
  select(-DEG_status,-time,-sample) %>%
  distinct() %>%
  column_to_rownames(var="target_id") %>%
  select(Guy11_0h, Guy11_4h, Guy11_24h,
         Dbip1_0h,Dbip1_4h,Dbip1_24h,
         Dmst12_0h,Dmst12_4h,Dmst12_24h)

downreg_matrix<-data.matrix(downreg)

# Use hierarchical clustering
mycols = colorRampPalette(c("blue","white","red"))(100) 
hr <- hclust(as.dist(1-cor(t(downreg_matrix), method="pearson")), method="average") 
dend <- as.dendrogram(hr) # Convert to dendrogram
row_order <- order.dendrogram(dend) # Row order from the hierarchical clustering
downreg_matrix_ordered <- downreg_matrix[row_order, ] %>%  # Reorder matrix based on the clustering
  as.data.frame() %>%
  rownames_to_column(var = "target_id")  # Add row names as a column

# Save ordered matrix
save_data(downreg_matrix_ordered, "downreg_matrix_ordered", format = "txt")

pdf_path <- here::here("data/results/00_figures/003_co_expression", "downregs_guy11vsDbip1_Dmst12.pdf")
#pdf(pdf_path, width = 12, height = 20)
heatmap.2(downreg_matrix, 
                  Rowv = as.dendrogram(hr), 
                  dendrogram = "row", 
                  trace = "none",  
                  col = mycols, 
                  keysize = 1, 
                  key.title = "Title", 
                  Colv = FALSE,
                  scale = c("row"), 
                  cexRow = 0.5, 
                  cexCol = 1, 
                  srtCol = NULL,  
                  srtRow = NULL, 
                  adjRow = c(0,NA), 
                  sepcolor = "white", 
                  margins = c(7,15), 
                  lwid = c(2,8),
                  lhei = c(2,8), 
                  main = "downregs", 
                  na.color = "black")
```

```{r}
sessionInfo()
```
